---
title: "Arrange Tables"
author: "Sabrina Hogan"
date: "`r doc_date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Intro to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Real life situation
I recently got asked to create a cohort of patients and collect samples from the biobank.  

Here are the selection criteria:  

- FFPE blocks
- in a window of 90days prior or after treatment initiation
- treatment of interest : immunotherapies


I had access to 2 tables different tables:  

- one containing the treatment information for some patients of interest
- one containing the sample information for all the patients of dermatology


# Explore the datasets

## Start your script
```{r, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}
rm(list=ls())

library(tidyverse)
library(lubridate)
```


## Import the datasets

```{r, message = FALSE, warning = FALSE, echo = TRUE, eval = TRUE}
samples <- read.delim("/data/Sabrina/2018_Cancer_immunotherapy/data/Sabrina_all_samples.txt", comment.char="#", na.strings = c("NA"," ",""))
treatments <- read.delim("/data/Sabrina/2018_Cancer_immunotherapy/data/Sabrina_treatments_table.txt", comment.char="#", na.strings = c("NA"," ",""))


tbl_df(samples)
str(samples)

tbl_df(treatments)
str(treatments)

```


## Rename
Some columns have weird names. Let's change this.  
Rename as follow:

- DBDEINDAT --> sampling.date
- DBCDIAGNOSE  --> diagnostic1
- DiagnoseExtractedFromDBBBLOB --> diagnostic2

```{r, message = TRUE, warning = FALSE, echo = TRUE, eval = TRUE}
# rename some columns
samples <- samples %>%
  rename(sampling.date=DBDEINDAT)%>%
  rename(diagnostic1=DBCDIAGNOSE)%>%
  rename(diagnostic2=DiagnoseExtractedFromDBBBLOB)
```

# Arrange tables

## treatment table

The treatment table contains information for a lot of different treatments.  
We need to select for immunotherapies only (aCLTA4, aPD1, aCTLA4_aPD1)


you should have:  

111 aCTLA4  
71 aCTLA4_aPD1  
195 aPD1   

you can use table() to explore the treatment colum.  


How do you solve this problem?

good. Now we can select our patients again.


*note: all the levels are still there even if we do not have any samples for those therapy anymore. It can sometimes also appear in your plot legend. This is how you would relevel the treatment column:*



## samples table

The sample table contains some empty rows which are obviously useless.  
Remove the rows which do not contain any information.


The sample table has multiple colums containing sampleID information.  
The usual nomenclature for our samples contains information about the type of sample, the year and the sample number, all together.

You need to combine  "sample.kind", "sample.year" and "sample.num" as follow:  

--> H12.345



Make sure there is no space between the characters...



# Select the right samples
Now that the nomenclature is correct, let's have a look at the different type of samples present.  
The table contains many sample types (HG=blood, M=mutation analysis etc...).  

Filter out the samples that are not parafin blocks (remove HG, M, Z, G, SP, TU, K)  


# Merge treatment and sample table

in order to plot the results we need all the information in the same table


# Select the samples in the correct period
subset the table in order to only keep the samples that have been collected between 90d before and 90d after treatment start.  




# Plot your results



what is wrong in the plot?  



now that your results look good, save your table.  

plot responders only and add a title containing the number of R and NR in the entire cohort  

